---
title: "GSE270486 Wounded dataset: putative CAM analysis"
author: "Michelle Meier"
date: "`r format(Sys.time(), '%d. %B %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

# Introduction

The purpose of this analysis is to investigate the putatitive CAM population Shimizu et al have presented in their Matter Arising to Nature. 

This is a dataset that we have made publicly available through our latest study [Convergent evolution of scavenger cell development at brain borders](https://www.nature.com/articles/s41586-025-10003-3) and we will be using the RDS object we provided for the Wounded 7dpf dataset (GSE270486).


Important dataset specs:

- Dev stage: 7dpf 

- Tissue: Head

- Sorting: kdrl, mCherry+ FACS (very leaky, lots of other cells)


# Set-up

## Load libraries

```{r}
suppressPackageStartupMessages({
  library(here)
  library(scran)
  library(scater)
  library(scuttle)
  library(scales)
  library(readr)
  library(Seurat)
  library(ggpubr)
  library(tidyverse)
  library(qs2)
  library(patchwork)
})

```


## Load data

```{r}
seurat <- qs_read(here("data/Level_01_D10026_Wounded.qs2"))

```

## Plotting themes

```{r}
plot_sizing <- theme_bw()+ theme(axis.title = element_text(size = 16),
        axis.text = element_text(size = 14, colour = "black"),
        plot.title = element_text(size = 18),
        legend.text = element_text(size = 14),
        legend.title = element_blank()) 


plot_sizing_seurat <- theme(axis.title = element_text(size = 16),
        axis.text = element_text(size = 14, colour = "black"),
        plot.title = element_text(size = 18),
        legend.text = element_text(size = 14))

rotate_x <- theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

strip_fix <- theme(strip.background = element_rect(fill = "white"), 
                   strip.text = element_text(size = 14))

```

## Colours 

```{r}
wounded_colours <- c(
  "#dbe2c6", #wildtype
  "#657c95" #wounded
)

names(wounded_colours) <- c("Control_kdrl", "Wounded_kdrl")

seurat$Sample_Name <- factor(seurat$Sample_Name, levels = names(wounded_colours)) #wildtype always first
```




# Overview 

## Cell type labels

Show cell type labels:

```{r, fig.align='center', fig.height=5, fig.width=8}

DimPlot(seurat, label = T,repel=T) + ggtitle("Wounded: cell type labels")

```

## TBI status for cell types

This dataset contains both wounded as well as control samples. Most of the immune cells actually come from the wounded sample.

```{r, fig.align='center'}

seurat@meta.data %>% 
  group_by(L1_Predicted_phenotype, Sample_Name) %>% 
  summarise(., count =n()) %>% 
  ggplot(., aes(x = L1_Predicted_phenotype, y= count, fill=Sample_Name)) +
  geom_bar(stat="identity", position = "fill") +
  scale_fill_manual(values = wounded_colours) +
  plot_sizing + rotate_x + xlab("") + 
  ggtitle("Wounded composition per celltype")

```
Get absolute numbers for macrophages. Neutrophils are all wounded 

```{r}

seurat@meta.data %>% 
  group_by(L1_Predicted_phenotype, Sample_Name) %>% 
  summarise(., count = n()) %>% 
  filter(., L1_Predicted_phenotype == "Macrophage")

```



# Subset to Immune cells

## Write function for overview

```{r}
expression_colours <- c("#d9d9d9", "#40004b")
colours_myeloid_celltypes <- c("grey", "black", "red")

make_wounded_subfigures <- function(object, genes_to_plot=genes){ #title
  p1 <- DimPlot(object, group.by = "myeloid_celltype", cols = colours_myeloid_celltypes) + NoAxes() + ggtitle("Myeloid populations (n=349)")
  p2 <- DimPlot(object, group.by = "Sample_Name", cols = wounded_colours) + NoAxes() + ggtitle("Wounding status")
  p3 <- DotPlot(object, features = genes_to_plot, group.by = "myeloid_celltype", cols = expression_colours) + rotate_x + ylab("") + ggtitle("Gene expression patterns") #change genes
  p4 <- object@meta.data %>% 
  group_by(myeloid_celltype, Sample_Name) %>% 
  summarise(., count =n()) %>% 
  ggplot(., aes(x = myeloid_celltype, y= count, fill=Sample_Name)) +
  geom_bar(stat="identity", position = "stack") + 
  scale_fill_manual(values = wounded_colours) +
  geom_text(aes(label = count), 
            position = position_stack(vjust=0.5),size=5, color="black",
            )+
  plot_sizing  + xlab("cluster")  +rotate_x + ggtitle("Wounded status composition")
  
  plot <- (p1+p2+plot_layout(ncol=2, axes="collect")) +p3+p4 & theme(plot.title = element_text(size = 18, face = "italic"))
  
  #  #add plotlayout
  return(plot)
  
  
}

```

## filter + reprocess

```{r}
seurat <- SetIdent(seurat, value = "L1_Predicted_phenotype")
myeloid <- subset(seurat, idents = c("Macrophage", "Neutrophil"))
myeloid

# Github issue [here](https://github.com/satijalab/seurat/issues/6719) suggests that rerunning SCTransform actually is not necessarily required
```
They don't say that they've rescaled/re-run SCT.

```{r}
myeloid <- myeloid %>% 
  RunPCA(., verbose = F) %>% 
  FindNeighbors(dims = c(1:10), verbose = F) %>% 
  FindClusters(resolution = 0.5, verbose = F) %>% 
  RunUMAP(dims = c(1:10), verbose = F)

```


Make the two plots to see if we can roughly get their clustering structure + gene expression patterns. 

```{r, fig.align='center', fig.width=12, fig.height=5}
umap <- DimPlot(myeloid)

#update genes
genes <- c("apoc1", "apoeb", "p2ry12", "sall3a", #microglia
           "csf1rb", "spi1a", "mpeg1.1", "marco", #shared
           "cybb", "f13a1b", "cd74b", "tgfbi", #CAM
            "lyve1a","lyve1b", "mrc1a","mrc1b" , "dab2", "stab1", "stab2" , #functional genes for BAM equivalence
           "mpx", "mmp9", "lyz"#neutro
           )

dotplot <- DotPlot(myeloid, features = genes, group.by = "SCT_snn_res.0.5") + rotate_x

umap + dotplot + plot_layout(widths = c(1,2))
```

Cluster 0,1 and 3 together would be microglia, cluster 2 would be putative "CAM" and cluster 4 neutrophils. 

```{r}
myeloid$SCT_snn_res.0.5 %>% table()
```


Adding cell type labels with n and making pretty plots. 

```{r}
myeloid$myeloid_celltype <- case_when(myeloid$SCT_snn_res.0.5 %in% c(0,1,3) ~ "MG\nn=252",
                                      myeloid$SCT_snn_res.0.5 == 2 ~ "MG/Macro subpop\nn=70",
                                      myeloid$SCT_snn_res.0.5 == 4 ~ "Neutrophil\nn=27")

```


```{r, fig.align='center', fig.height=8, fig.width=15, warning=FALSE}
names(colours_myeloid_celltypes) <- c("MG\nn=252", "Neutrophil\nn=27", "MG/Macro subpop\nn=70")

make_wounded_subfigures(myeloid) + plot_annotation(title = "Wounded dataset 7dpf (GSE270486)")
```


# For figure: individual panels

Make individual panels for illustrator figure. This ensures we keep the same dimensions. We also tend to adjust things like text size in illustrator as its a bit easier. 

## Functions 

Write some core functions. 

```{r}
plot_umap <- function(seurat.object = myeloid,
                      group, 
                      title,
                      save.dir = output,
                      cols = NULL,
                      project.name = name, 
                      size = 10){
  #with legend
  umap <- DimPlot(seurat.object,
                  group.by = group,
                  cols = cols,
                  shuffle = T, 
                  pt.size = 1,
                  label = F) + NoAxes() + ggtitle(title)
  
  umapFileName <- sprintf('%s/%s_umap_%s.pdf', save.dir, project.name, group)
  ggsave(umapFileName,
         umap,
         device="pdf",
         width=size,
         height=size,
         units="cm",
         scale=1)
  
  #without legend
  umap.nolegend <- umap + NoLegend() + NoAxes() + ggtitle('')
  umapFileName <- sprintf('%s/%s_umap_%s_nolegend.pdf', save.dir, project.name, group)
  ggsave(umapFileName,
         umap.nolegend,
         device="pdf",
         width=size,
         height=size,
         units="cm",
         scale=1)
  
  return(umap)
}


```

## Make panels

```{r}
output <- here("output/figures/")
name <- "Wounded_GSE270486"
```


```{r, fig.align='center', fig.height=8, fig.width=15, warning=FALSE}

#use function
p1 <- plot_umap(group = "myeloid_celltype", cols = colours_myeloid_celltypes, title = "Myeloid populations (n=349)")
p2 <- plot_umap(group = "Sample_Name", cols = wounded_colours, title = "Wounding status")

#dotplot
p3 <- DotPlot(myeloid, features = genes, group.by = "myeloid_celltype", cols = expression_colours) + rotate_x + ylab("") + ggtitle("Gene expression patterns") 
ggsave(sprintf("%s/%s_dotplot_myeloid_celltype.pdf", output, name),
         p3,
         device="pdf",
         width=20,
         height=10,
         units="cm",
         scale=1)

#stacked barplot
p4 <- myeloid@meta.data %>% 
  group_by(myeloid_celltype, Sample_Name) %>% 
  summarise(., count =n()) %>% 
  ggplot(., aes(x = myeloid_celltype, y= count, fill=Sample_Name)) +
  geom_bar(stat="identity", position = "stack") + 
  scale_fill_manual(values = wounded_colours) +
  geom_text(aes(label = count), 
            position = position_stack(vjust=0.5),size=5, color="black")+
  plot_sizing  + xlab("")  +rotate_x + ggtitle("Wounded status composition")
ggsave(sprintf("%s/%s_stacked_barplot_myeloid_celltype_wounded.pdf", output, name),
         p4,
         device="pdf",
         width=13,
         height=10,
         units="cm",
         scale=1)

  
(p1+p2+plot_layout(ncol=2, axes="collect")) +p3+p4 
```






