---
title: "GSE270486/GSE188342 Timecourse dataset: make Level 02 (immune cell) object"
author: "Michelle Meier"
date: "`r format(Sys.time(), '%d. %B %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

# Introduction

The purpose of this analysis is to investigate the putative CAM population Shimizu et al have presented in their Matter Arising to Nature. Specifically, this looks at the timecourse dataset, spanning 40hpf - 30dpf. 


The first step for pre-processing can be found [here](Gaudi_Timecourse_40hpf_30dpf_makeObject_Level_01_allCells.html). In this markdown, I will subset and reprocess to only include immune cells. I have selected the following genes, which roughly correspond to the genes used in Shimizu et al. 

```{r}
genes_immune_select <- c("apoc1", "apoeb", "p2ry12", #microglia
                          "tgfbi","f13a1b", "cybb" , #putative CAM
                         "lyz", "mpx", "mmp9",  #neutrophil
                         "runx3", "skap1", "gata3", #lymphocytes
                         "mpeg1.1" #MG + Macro
                         )
```



## Load libraries


```{r}
suppressPackageStartupMessages({
  library(here)
  library(scran)
  library(scater)
  library(scuttle)
  library(scales)
  library(readr)
  library(Seurat)
  library(ggpubr)
  library(tidyverse)
  library(qs2)
  library(patchwork)
  library(simspec)
  library(scDblFinder)
  library(clustree)
})

```


## load data


```{r}

Level_01_seuratObject <- qs_read(here("output/data/seurat_objects/Timecourse_40hpf_30dpf_Level_01.qs2"))

Level_02_seuratObject_old <- qs_read("/Users/meiermichelle/Library/CloudStorage/OneDrive-PeterMac/HOGAN_LAB/analysis/BAM_quest/output/data/seurat_objects/Timecourse_40hpf_30dpf_Level_02.qs2")



```

## Plotting themes

```{r}
plot_sizing <- theme_bw()+ theme(axis.title = element_text(size = 16),
        axis.text = element_text(size = 14, colour = "black"),
        plot.title = element_text(size = 18),
        legend.text = element_text(size = 14),
        legend.title = element_blank()) 


plot_sizing_seurat <- theme(axis.title = element_text(size = 16),
        axis.text = element_text(size = 14, colour = "black"),
        plot.title = element_text(size = 18),
        legend.text = element_text(size = 14))

rotate_x <- theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

strip_fix <- theme(strip.background = element_rect(fill = "white"), 
                   strip.text = element_text(size = 14))

```


## Colours 

```{r}
#celltype colours

celltype_colours <- c("red2", "black", "grey")
names(celltype_colours) <- c("MG/Macro subpop", "Neutrophil", "MG")

stage.colours <- c( '#e4c6d3','#be7794', '#a95175', '#823e59', '#3f1e2c', "#190c11")
names(stage.colours) <-  c('40hpf', '3dpf', '4dpf', '5dpf', '7dpf', "30dpf")

expression_colours <-  c("#d9d9d9", "#7a0177")

```


# Identify population of interest

Use the genes we decided on. Make UMAPs first.

```{r, fig.align='center', fig.height=15, fig.width=10}
FeaturePlot(object = Level_01_seuratObject, features = genes_immune_select)

```


Make a dotplot, but first look at some resolutions. 

```{r, fig.align='center', fig.height=8, fig.width=10}
clustree(Level_01_seuratObject)
```

Visualise clusters from res 0.3-0.6.

```{r, fig.align='center', fig.height=10, fig.width=10}
resolutions <- paste0("RNA_snn_res.", seq(0.3, 0.6, 0.1))


DimPlot(object = Level_01_seuratObject, group.by = resolutions, shuffle = T)
```

None of these really make a difference for the clusters I think are immune cells. Pick 0.3

```{r, fig.align='center', fig.height=5, fig.width=12}

p1 <- DotPlot(object = Level_01_seuratObject, features = genes_immune_select, group.by = "RNA_snn_res.0.3") + rotate_x + ylab("") + xlab("")
p2 <- DimPlot(object = Level_01_seuratObject, group.by = "RNA_snn_res.0.3", shuffle = T, label = T) + theme(legend.position = "none")
p3 <- DimPlot(object = Level_01_seuratObject, group.by = "scDblFinder.class", shuffle = T) 

p1+p2+p3
```


Cluster 14 seems to be a doublet cluster. Cluster 15 seems to express the microglia genes, but is pretty far away from all the other clusters (but UMAP distance is not meaningful, so who knows)


Get cluster metrics.

```{r, fig.align='center', fig.height=8, fig.width=10}
plot_list <- list()

plot_list[[1]] <- Level_01_seuratObject@meta.data %>% 
  ggplot(., aes(y = log10(nCount_RNA), x = RNA_snn_res.0.3, fill=RNA_snn_res.0.3)) +
  geom_boxplot() +
  plot_sizing + theme(legend.position = "none") + ggtitle("Library size")

plot_list[[2]] <- Level_01_seuratObject@meta.data %>% 
  ggplot(., aes(y = log10(nFeature_RNA), x = RNA_snn_res.0.3, fill=RNA_snn_res.0.3)) +
  geom_boxplot() +
  plot_sizing + theme(legend.position = "none") + ggtitle("#genes")

plot_list[[3]] <- Level_01_seuratObject@meta.data %>% 
  ggplot(., aes(y = percent.mt, x = RNA_snn_res.0.3, fill=RNA_snn_res.0.3)) +
  geom_boxplot() +
  plot_sizing + theme(legend.position = "none") + ggtitle("%mito")

wrap_plots(plot_list, axes = "collect", ncol=2)
```

cluster 12 (likely macrophages) higher %mito, active cells

# Re-process immune cells 

## Subset

```{r}
Level_01_seuratObject <- SetIdent(Level_01_seuratObject, value = "RNA_snn_res.0.3")
Level_02_seuratObject <- subset(Level_01_seuratObject, idents = c(9,12))
Level_02_seuratObject
```


## reprocess


Reprocess with the same steps, including css.

```{r}
Level_02_seuratObject	<- NormalizeData(Level_02_seuratObject, verbose = F) %>% #shouldn't be needed but won't hurt
  FindVariableFeatures(. , verbose = F) %>%
  ScaleData(., split.by = "SampleName",features = rownames(Level_02_seuratObject), verbose = F) %>% #using split by argument because we have massive library size effects
  RunPCA(.,npcs = 20, verbose = F) 

Level_02_seuratObject <- cluster_sim_spectrum(object = Level_02_seuratObject, 
                                              label_tag = "SampleName") 

Level_02_seuratObject <-  RunUMAP(Level_02_seuratObject,
                                  reduction = "css",
                                  dims = 1:ncol(Embeddings(Level_02_seuratObject, "css")), verbose = F) %>% 
  FindNeighbors(.,reduction = "css",
                dims = 1:ncol(Embeddings(Level_02_seuratObject, "css")), verbose = F) %>% 
  FindClusters(., resolution = seq(0.1, 1, 0.1), verbose = F)

```



## UMAP overviews

```{r, fig.align='center', fig.height=10, fig.width=15}
plot_list <- list()

plot_list[[1]] <- FeaturePlot(object = Level_02_seuratObject, features = "nCount_RNA", min.cutoff = "q1", max.cutoff = "q99")
plot_list[[2]] <- FeaturePlot(object = Level_02_seuratObject, features = "nFeature_RNA", min.cutoff = "q1", max.cutoff = "q99")
plot_list[[3]] <- FeaturePlot(object = Level_02_seuratObject, features = "percent.mt")
plot_list[[4]] <- DimPlot(object = Level_02_seuratObject, group.by = "SampleName", shuffle = T)
plot_list[[5]] <- DimPlot(object = Level_02_seuratObject, group.by = "Stage", shuffle = T)
plot_list[[6]] <- DimPlot(object = Level_02_seuratObject, group.by = "scDblFinder.class", shuffle = T)


wrap_plots(plot_list, axes = "collect")

```

## Clusters

```{r,fig.align='center', fig.height=5, fig.width=5}
DimPlot(object = Level_02_seuratObject, group.by = "RNA_snn_res.0.1", label = T) + theme(legend.position = "none") 
```

Boxplots with metrics

```{r, fig.align='center', fig.height=10, fig.width=10}
plot_list <- list()

plot_list[[1]] <- Level_02_seuratObject@meta.data %>% 
  ggplot(., aes(y = log10(nCount_RNA), x = RNA_snn_res.0.1, fill=RNA_snn_res.0.1)) +
  geom_boxplot() +
  plot_sizing + theme(legend.position = "none") + ggtitle("Library size")

plot_list[[2]] <- Level_02_seuratObject@meta.data %>% 
  ggplot(., aes(y = log10(nFeature_RNA), x = RNA_snn_res.0.1, fill=RNA_snn_res.0.1)) +
  geom_boxplot() +
  plot_sizing + theme(legend.position = "none") + ggtitle("#genes")

plot_list[[3]] <- Level_02_seuratObject@meta.data %>% 
  ggplot(., aes(y = percent.mt, x = RNA_snn_res.0.1, fill=RNA_snn_res.0.1)) +
  geom_boxplot() +
  plot_sizing + theme(legend.position = "none") + ggtitle("%mito")

wrap_plots(plot_list, axes = "collect", ncol=2)
```


# Annotation

## Gene expression UMAPs


```{r, fig.align='center', fig.height=20, fig.width=20}
FeaturePlot(object = Level_02_seuratObject, features = c(genes_immune_select, "rna_mki67"))
```

Make dotplot

```{r, fig.align='center', fig.height=5, fig.width=10}
p1 <- DotPlot(object = Level_02_seuratObject, features = genes_immune_select, group.by = "RNA_snn_res.0.1")+ rotate_x
p2 <- DimPlot(object = Level_02_seuratObject, group.by = "RNA_snn_res.0.1", label = T) +theme(legend.position = "none")

p1+p2
```

Add annotation based on gene expression pattern


```{r}
Level_02_seuratObject$prelim_celltype <- case_when(Level_02_seuratObject$RNA_snn_res.0.1 == 1 ~ "MG/Macro subpop",
                                                   Level_02_seuratObject$RNA_snn_res.0.1 == 3 ~ "MG",
                                                   Level_02_seuratObject$RNA_snn_res.0.1 %in% c(0,2) ~ "Neutrophil")

```

```{r}
Level_02_seuratObject$prelim_celltype %>% table()
```

Add fancy names with name with N. 

```{r}
Level_02_seuratObject$prelim_celltype <- case_when(Level_02_seuratObject$RNA_snn_res.0.1 == 1 ~ "MG/Macro subpop\nn=941",
                                                   Level_02_seuratObject$RNA_snn_res.0.1 == 3 ~ "MG\nn=133",
                                                   Level_02_seuratObject$RNA_snn_res.0.1 %in% c(0,2) ~ "Neutrophil\nn=1651")

Level_02_seuratObject$prelim_celltype <- as.factor(Level_02_seuratObject$prelim_celltype)
```

cell type vs stage.

```{r}
table(Level_02_seuratObject$prelim_celltype, Level_02_seuratObject$Stage)
```


# Export 

Let's save this! 

```{r}
filename <- here("output/data/seurat_objects/Timecourse_40hpf_30dpf_Level_02_immuneCells.qs2")
qs_save(object = Level_02_seuratObject, file = filename)
```



