---
title: "GSE270486/GSE188342 Timecourse dataset: make Level 01 (all cells) object"
author: "Michelle Meier"
date: "`r format(Sys.time(), '%d. %B %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

# Introduction

The purpose of this analysis is to investigate the putatitive CAM population Shimizu et al have presented in their Matter Arising to Nature. Specifically, this looks at the timecourse dataset, spanning 40hpf - 30dpf. 

All the data used for this analysis are publicly available under either [GSE270486](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE270486) or [GSE188342](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE188342). The exception to this is the CellBender corrected count matrix for the 30dpf sample ([GSM9038467](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM9038467)) - we will be making these available on zenodo. 


Its important to note that the different samples were processed in different ways. The table bellow provides an overview. Also, we are only including wildtype samples - the 7dpf wounded as well as the 30dpf osr2-/- sample are not included. 

Important sample specs:

|**Dev Stage**|**Samples**|**Tissue**|**Sorting**|
|:---------:|:---------:|:---------:|:---------:|
|40hpf|Sample 20043, Sample 20048, Sample 20049|**Whole embryo**|fli1a|
|3dpf|Sample 20044|**Whole embryo**|lyve1b, kdrl|
|4dpf|Sample 20125|**Whole embryo**|fli1a, lyve1b|
|5dpf|Sample 20045|**Whole embryo**|lyve1b, kdrl|
|7dpf|Sample 20129|**Head**|kdrl|
|30dpf|Sample 20199|**Head**|fli1a, lyve1b|



## Load libraries


```{r, warning=FALSE}
suppressPackageStartupMessages({
  library(here)
  library(scran)
  library(scater)
  library(scuttle)
  library(scales)
  library(readr)
  library(Seurat)
  library(ggpubr)
  library(tidyverse)
  library(qs2)
  library(patchwork)
  library(simspec)
  library(scDblFinder)
})

```


## get paths

```{r}
ribo_genes <- read.table(here("data/DR_ribosomal_globin_genes.csv"),
													sep="\t",
													stringsAsFactors=FALSE) %>% 
  pull(gene_name) %>% as.character()
  

sample_20043 <- here("data/GeneralCellrangerOutputs/Sample_20043/filtered_feature_bc_matrix")
sample_20044 <- here("data/GeneralCellrangerOutputs/Sample_20044/filtered_feature_bc_matrix")
sample_20045 <- here("data/GeneralCellrangerOutputs/Sample_20045/filtered_feature_bc_matrix")
sample_20048 <- here("data/GeneralCellrangerOutputs/Sample_20048/filtered_feature_bc_matrix")
sample_20049 <- here("data/GeneralCellrangerOutputs/Sample_20049/filtered_feature_bc_matrix")
sample_20125 <- here("data/GeneralCellrangerOutputs/Sample_20125/filtered_feature_bc_matrix")
sample_20129 <- here("data/GeneralCellrangerOutputs/Sample_20129/filtered_feature_bc_matrix")
sample_20199 <- here("data/GeneralCellrangerOutputs/S10199_cellbender_output_filtered_seurat.h5") #cellbender


```

## Plotting themes

```{r}
plot_sizing <- theme_bw()+ theme(axis.title = element_text(size = 16),
        axis.text = element_text(size = 14, colour = "black"),
        plot.title = element_text(size = 18),
        legend.text = element_text(size = 14),
        legend.title = element_blank()) 


plot_sizing_seurat <- theme(axis.title = element_text(size = 16),
        axis.text = element_text(size = 14, colour = "black"),
        plot.title = element_text(size = 18),
        legend.text = element_text(size = 14))

rotate_x <- theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

strip_fix <- theme(strip.background = element_rect(fill = "white"), 
                   strip.text = element_text(size = 14))

```


# Generate seurat object

Write function to read in data for cellranger outputs.

```{r}

read_cellranger <- function(path,
                            genes_remove = ribo_genes,
                            sample_id, 
                            stage){
  counts <- Read10X(path)
  filtered_counts <- counts[!(rownames(counts) %in% genes_remove),] #remove ribo genes
  
  seuratObject 						<- CreateSeuratObject(
													counts=filtered_counts,
													min.cells = 5)
  
  seuratObject$SampleName <- sample_id
  seuratObject$Stage <- stage
  
  return(seuratObject)
  
}

```

Let's go.

```{r}
list_seurat <- list()
```

```{r}
list_seurat[["sample_20043"]] <- read_cellranger(path = sample_20043, stage = "40hpf", sample_id = "sample_20043")
list_seurat[["sample_20044"]] <- read_cellranger(path = sample_20044, stage = "3dpf", sample_id = "sample_20044")
list_seurat[["sample_20045"]] <- read_cellranger(path = sample_20045, stage = "5dpf", sample_id = "sample_20045")
list_seurat[["sample_20048"]] <- read_cellranger(path = sample_20048, stage = "40hpf", sample_id = "sample_20048")
list_seurat[["sample_20049"]] <- read_cellranger(path = sample_20049, stage = "40hpf", sample_id = "sample_20049")
list_seurat[["sample_20125"]] <- read_cellranger(path = sample_20125, stage = "4dpf", sample_id = "sample_20125")
list_seurat[["sample_20129"]] <- read_cellranger(path = sample_20129, stage = "7dpf", sample_id = "sample_20129")

```

Add cellbender sample (30dpf) as well.

```{r}
counts <- Read10X_h5(sample_20199)
seurat	<- CreateSeuratObject(
													counts=counts[!(rownames(counts) %in% ribo_genes),],
													min.cells = 5)
seurat$SampleName <- "sample_20199"
seurat$Stage <- "30dpf"

list_seurat[["sample_20199"]] <- seurat


#clean up 
rm(counts, seurat);gc()

```



Add % mito to all objects.

```{r}
list_seurat <- lapply(list_seurat, function(seurat){
  seurat$`percent.mt` <- PercentageFeatureSet(seurat,
                                                pattern = "^[mtMtMT-]")
  return(seurat)
})

```



# QC 

Make function for densitiy and correlation plots. 

```{r}
make_density_plots <- function(seurat){
  plot_list <- list()

  #ncount
  p1 <- ggplot(seurat@meta.data, aes(x=log10(nCount_RNA))) + 
    geom_histogram(aes(y=..density..), colour="black", fill="white")+
    geom_density(alpha=.2, fill="#FF6666")+ xlab('log10(count)') +
    plot_sizing + ggtitle(sprintf("Library size, %s", seurat$SampleName))
  
  #nfeat
  p2 <- ggplot(seurat@meta.data, aes(x=log10(nFeature_RNA))) + 
    geom_histogram(aes(y=..density..), colour="black", fill="white")+
    geom_density(alpha=.2, fill="#FF6666")+ xlab('log10(nfeature)') +
    plot_sizing + ggtitle(sprintf("#features, %s",  seurat$SampleName))
  
  #mito
  p3 <- ggplot(seurat@meta.data, aes(x=percent.mt)) + 
    geom_histogram(aes(y=..density..), colour="black", fill="white")+
    geom_density(alpha=.2, fill="#FF6666")+ xlab('percent.mt') +
    plot_sizing + ggtitle(sprintf("mito, %s",  seurat$SampleName))
  
  return(p1+p2+p3)
  
}

make_correlation_plots <- function(seurat){
  plot_list <- list()

  meta <- seurat@meta.data %>% 
    mutate(., log10ncount = log10(nCount_RNA),
              log10nfeat = log10(nFeature_RNA))
  
  #ncount
  p1 <- ggscatter(meta, x = "log10ncount", y = "log10nfeat",
          add = "reg.line",                                 
          conf.int = TRUE,                                
          add.params = list(color = "red4",
                            fill = "lightgray")) +
    plot_sizing + ylab("log10(#features)") + xlab("log10(nCount)") +
    stat_cor(method = "pearson", label.x = 2.5, label.y = 4, size = 5) +
    ggtitle(sprintf("Lib size vs #features, %s", seurat$SampleName))
  
  #nfeat
  p2 <- ggscatter(meta, x = "log10ncount", y = "percent.mt",
          add = "reg.line",                                 
          conf.int = TRUE,                                
          add.params = list(color = "red4",
                            fill = "lightgray")) +
    plot_sizing + ylab("%mito") + xlab("log10(nCount)") +
    stat_cor(method = "pearson", label.x = 2.5, label.y = 80, size = 5) +
    ggtitle(sprintf("Lib size vs mito, %s",  seurat$SampleName))
  
  #mito
  p3 <- ggscatter(meta, x = "log10nfeat", y = "percent.mt",
          add = "reg.line",                                 
          conf.int = TRUE,                                
          add.params = list(color = "red4",
                            fill = "lightgray")) +
    plot_sizing + ylab("%mito") + xlab("log10(#features)") +
    stat_cor(method = "pearson", label.x = 2.5, label.y = 80, size = 5) +
    ggtitle(sprintf("#features vs mito, %s",  seurat$SampleName))
  
  return(p1+p2+p3)
  
}
```


## Lib size, nfeatures, %mito


```{r, fig.align='center', fig.height=30, fig.width=10, warning=FALSE}
plot_list <- lapply(list_seurat, make_density_plots)

wrap_plots(plot_list, ncol=1) + plot_layout(axis_titles = "collect")
```

We know from previous explorations that the osr2 experiment did a weird thing where we get two peaks. We will have to manually filter that one. All other samples can be filtered as done previously. 


## Correlations

```{r, fig.align='center', fig.height=30, fig.width=15, warning=FALSE}

plot_list <- lapply(list_seurat, make_correlation_plots)

wrap_plots(plot_list, ncol=1) + plot_layout(axis_titles = "collect")

```


# Doublet calling

We do not usually remove doublets for developmental datasets as we've observed that this step removes true biological intermediates. We'll predict doublets, and check how this aligns with 10X expected doublet values.


```{r}
run_scDblFinder <- function(seurat){
  sce <- as.SingleCellExperiment(seurat) #convert
  
  sce <- logNormCounts(sce) %>% #run steps
  runPCA() %>%
  runUMAP()
  
  set.seed(1997)#for reproducibility
  sce <- scDblFinder(sce, clusters = T) #run scDblFinder with clusters,verbose=F
  
  #add to seurat
  seurat$scDblFinder.class <- sce$scDblFinder.class #actual label
  seurat$scDblFinder.cxds_score <- sce$scDblFinder.cxds_score #co-expression score
  
  return(seurat)
}
```


```{r}
list_seurat <- lapply(list_seurat, run_scDblFinder)
```


Compute expected vs detected doublet rate per sample. All of these should be 10x, so they follow the 8% per 10k. 

```{r}
comp_table <- lapply(list_seurat, function(seurat){
  df_out <- data.frame(expected = (ncol(seurat)* 0.008)/1000,
                       observed = nrow(seurat@meta.data %>% filter(scDblFinder.class == "doublet"))/ncol(seurat))
  return(df_out)
}) %>% bind_rows(., .id="sample")
```

```{r}
DT::datatable(comp_table)
```

You can clearly see that for the dev samples expected < observed - importantly, scDblFinder uses the expected value as prior, so there are definitively too many doublets detected for these samples. Lets not remove them, to preserve real intermediate stages. For the osr2 dataset, I suggest we remove because the metrics match + they loaded a lot of cells (20k) + they should not have any non-doublet intermediate stages anymore.

# Filter cells 

Make function to remove based on thresholds. I'm putting in the general lab threshold we've used which work well for us. In most cases...

```{r}
run_qc <- function(seurat, 
                   nmads=5,
                   nfeatures=200,
                   mt_threshold=20,
                   doublet_filter=FALSE){
  
  #detect nmads outliers
  lib_size_outliers 					<- isOutlier(seurat$nCount_RNA,nmads = nmads, log = T)

  seurat_filtered 						<- seurat[,!lib_size_outliers]

  #remove nFeature threshold
  seurat_filtered	<- subset(seurat_filtered,
												subset = nFeature_RNA > nfeatures)

  #	Remove mitochondria-high cells
  seurat_filtered 						<- subset(seurat_filtered,
												subset = percent.mt < mt_threshold)

  #	Remove doublets
  if (doublet_filter){
    seurat_filtered 						<- subset(seurat_filtered,
												subset = scDblFinder.class == "singlet")
  }
  
  
  return(seurat_filtered)
  
}
```


Run filter for each sample. 

```{r}
list_seurat[["sample_20043"]] <- run_qc(seurat = list_seurat[["sample_20043"]])
list_seurat[["sample_20044"]] <- run_qc(seurat = list_seurat[["sample_20044"]])
list_seurat[["sample_20045"]] <- run_qc(seurat = list_seurat[["sample_20045"]])
list_seurat[["sample_20048"]] <- run_qc(seurat = list_seurat[["sample_20048"]])
list_seurat[["sample_20049"]] <- run_qc(seurat = list_seurat[["sample_20049"]])
list_seurat[["sample_20125"]] <- run_qc(seurat = list_seurat[["sample_20125"]])
list_seurat[["sample_20129"]] <- run_qc(seurat = list_seurat[["sample_20129"]])

#different choices for osr2 dataset
list_seurat[["sample_20199"]] <- run_qc(seurat = list_seurat[["sample_20199"]], nfeatures = 600, doublet_filter = TRUE)

```

## post filter histograms

```{r, fig.align='center', fig.height=30, fig.width=10, warning=FALSE}
plot_list <- lapply(list_seurat, make_density_plots)

wrap_plots(plot_list, ncol=1) + plot_annotation(title = "post filtering")
```

# Merge + dimensionality reduction 

```{r}
Level_01_seuratObject <- merge(list_seurat[[1]],
                                     y = c(list_seurat[[2]],
                                           list_seurat[[3]],
                                           list_seurat[[4]],
                                           list_seurat[[5]],
                                           list_seurat[[6]],
                                           list_seurat[[7]],
                                          list_seurat[[8]]),
                                     project = "Timecourse", 
                                      merge.data = TRUE)

Level_01_seuratObject<- JoinLayers(Level_01_seuratObject) #seurat v5 things

Level_01_seuratObject
```


```{r}
#clean up
rm(list_seurat, comp_table, plot_list);gc()
```

## run dim reduction

Need to run this before integration (see their [github](https://github.com/quadbio/simspec)). 

```{r}
Level_01_seuratObject	<- NormalizeData(Level_01_seuratObject, verbose = F) %>% 
  FindVariableFeatures(. , verbose = F) %>% 
  ScaleData(., split.by = "SampleName",features = rownames(Level_01_seuratObject), verbose = F) %>% #using split by argument because we have massive library size effects
  RunPCA(.,npcs = 20, verbose = F) 

```


# Integrate with CSS (simspec)

We find this works well for our developmental datasets. Their [github](https://github.com/quadbio/simspec)

```{r}
#	Integrate data
Level_01_seuratObject 		<- cluster_sim_spectrum(
  object = Level_01_seuratObject,
  label_tag = "Stage")

Level_01_seuratObject 		<- RunUMAP(Level_01_seuratObject,
                                             reduction = "css",
                                             dims = 1:ncol(Embeddings(Level_01_seuratObject, "css")), verbose = F) %>% 
  FindNeighbors(.,reduction = "css",
                dims = 1:ncol(Embeddings(Level_01_seuratObject, "css")), verbose = F) %>% 
    FindClusters(., resolution = seq(0.1, 1, 0.1), verbose = F)
```

# Look at dataset 

## CSS correlations

Look at correlation of certain dataset metrics with CSS

```{r}
run_css_correlations <- function(seurat){
  plot_list <- list()

  meta <- seurat@meta.data %>% 
    mutate(., log10ncount = log10(nCount_RNA),
              log10nfeat = log10(nFeature_RNA)) %>% 
    select(., log10ncount, log10nfeat) %>% 
    rownames_to_column("Barcode") %>% 
    full_join(., seurat@reductions$css@cell.embeddings %>% as.data.frame() %>% rownames_to_column("Barcode") %>% select(., Barcode, CSS_1, CSS_2,CSS_3))
  
  p1 <- ggscatter(meta, x = "CSS_1", y = "log10ncount",
          add = "reg.line",                                 
          conf.int = TRUE,                                
          add.params = list(color = "red4",
                            fill = "lightgray")) +
    plot_sizing + ylab("log10(nCount)") + xlab("CSS_1") +
    stat_cor(method = "pearson", size = 5, colour = "red4") +
    ggtitle(sprintf("CSS_1 vs library size, %s", seurat@project.name))
  
  p2 <- ggscatter(meta, x = "CSS_2", y = "log10ncount",
          add = "reg.line",                                 
          conf.int = TRUE,                                
          add.params = list(color = "red4",
                            fill = "lightgray")) +
    plot_sizing + ylab("log10(nCount)") + xlab("CSS_2") +
    stat_cor(method = "pearson", size = 5, colour = "red4") +
    ggtitle(sprintf("CSS_2 vs library size, %s", seurat@project.name))
  
  p3<- ggscatter(meta, x = "CSS_3", y = "log10ncount",
          add = "reg.line",                                 
          conf.int = TRUE,                                
          add.params = list(color = "red4",
                            fill = "lightgray")) +
    plot_sizing + ylab("log10(nCount)") + xlab("CSS_3") +
    stat_cor(method = "pearson", size = 5, colour = "red4") +
    ggtitle(sprintf("CSS_3 vs library size, %s", seurat@project.name))
  
  return(p1+p2+p3)
}
```


```{r, fig.align='center', fig.height=5, fig.width=15}
run_css_correlations(Level_01_seuratObject)
```

## UMAP overviews


```{r, fig.align='center', fig.height=10, fig.width=15}
plot_list <- list()

plot_list[[1]] <- FeaturePlot(object = Level_01_seuratObject, features = "nCount_RNA", min.cutoff = "q1", max.cutoff = "q99")
plot_list[[2]] <- FeaturePlot(object = Level_01_seuratObject, features = "nFeature_RNA", min.cutoff = "q1", max.cutoff = "q99")
plot_list[[3]] <- FeaturePlot(object = Level_01_seuratObject, features = "percent.mt")
plot_list[[4]] <- DimPlot(object = Level_01_seuratObject, group.by = "SampleName", shuffle = T)
plot_list[[5]] <- DimPlot(object = Level_01_seuratObject, group.by = "Stage", shuffle = T)
plot_list[[6]] <- DimPlot(object = Level_01_seuratObject, group.by = "scDblFinder.class", shuffle = T)

wrap_plots(plot_list, axes = "collect")

```


Stage/Sample split - are the 40hpf sitting together?

```{r, fig.align='center', fig.height=5, fig.width=20}
DimPlot(object = Level_01_seuratObject, group.by = "SampleName", shuffle = T, split.by = "Stage")
```


## Per cluster metrics

```{r, fig.align='center', fig.height=10, fig.width=15}
plot_list <- list()

plot_list[[1]] <- Level_01_seuratObject@meta.data %>% 
  ggplot(., aes(y = log10(nCount_RNA), x = seurat_clusters, fill=seurat_clusters)) +
  geom_boxplot() +
  plot_sizing + theme(legend.position = "none") + ggtitle("Library size")

plot_list[[2]] <- Level_01_seuratObject@meta.data %>% 
  ggplot(., aes(y = log10(nFeature_RNA), x = seurat_clusters, fill=seurat_clusters)) +
  geom_boxplot() +
  plot_sizing + theme(legend.position = "none") + ggtitle("#genes")

plot_list[[3]] <- Level_01_seuratObject@meta.data %>% 
  ggplot(., aes(y = percent.mt, x = seurat_clusters, fill=seurat_clusters)) +
  geom_boxplot() +
  plot_sizing + theme(legend.position = "none") + ggtitle("%mito")

wrap_plots(plot_list, axes = "collect", ncol=2)
```

# Clean up

Shouldn't be much, but rearrange time points and make factors. 

```{r}
Level_01_seuratObject$Stage <- factor(Level_01_seuratObject$Stage, levels = c("40hpf", "3dpf", "4dpf" , "5dpf", "7dpf",  "30dpf"))
```


# Export 

Let's save this! 

```{r}
filename <- here("output/data/seurat_objects/Timecourse_40hpf_30dpf_Level_01.qs2")
qs_save(object = Level_01_seuratObject, file = filename)
```

