---
title: "Make Figures for putative CAM analysis"
author: "Michelle Meier"
date: "`r format(Sys.time(), '%d. %B %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

# Introduction

This markdown generates individual panels for illustrator figure. This ensures we keep the same dimensions. We also tend to adjust things like text size in illustrator as its a bit easier. 

## Load libraries


```{r}
suppressPackageStartupMessages({
  library(here)
  library(scran)
  library(scater)
  library(scuttle)
  library(scales)
  library(readr)
  library(Seurat)
  library(ggpubr)
  library(tidyverse)
  library(qs2)
  library(patchwork)
  library(simspec)
  library(scDblFinder)
  library(clustree)
})

```

## Functions 

Write some core functions. 

```{r}
plot_umap <- function(seurat.object,
                      group, 
                      title,
                      save.dir = output,
                      cols = NULL,
                      project.name, 
                      size = 10){
  #with legend
  umap <- DimPlot(seurat.object,
                  group.by = group,
                  cols = cols,
                  shuffle = T, 
                  pt.size = 1,
                  label = F) + NoAxes() + ggtitle(title)
  
  umapFileName <- sprintf('%s/%s_umap_%s.pdf', save.dir, project.name, group)
  ggsave(umapFileName,
         umap,
         device="pdf",
         width=size,
         height=size,
         units="cm",
         scale=1)
  
  #without legend
  umap.nolegend <- umap + NoLegend() + NoAxes() + ggtitle('')
  umapFileName <- sprintf('%s/%s_umap_%s_nolegend.pdf', save.dir, project.name, group)
  ggsave(umapFileName,
         umap.nolegend,
         device="pdf",
         width=size,
         height=size,
         units="cm",
         scale=1)
  
  return(umap)
}


```


## Plotting themes

```{r}
plot_sizing <- theme_bw()+ theme(axis.title = element_text(size = 16),
        axis.text = element_text(size = 14, colour = "black"),
        plot.title = element_text(size = 18),
        legend.text = element_text(size = 14),
        legend.title = element_blank()) 


plot_sizing_seurat <- theme(axis.title = element_text(size = 16),
        axis.text = element_text(size = 14, colour = "black"),
        plot.title = element_text(size = 18),
        legend.text = element_text(size = 14))

rotate_x <- theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

strip_fix <- theme(strip.background = element_rect(fill = "white"), 
                   strip.text = element_text(size = 14))

```


## Load data

```{r}
timecourse <- qs_read(here("output/data/seurat_objects/Timecourse_40hpf_30dpf_Level_02_immuneCells.qs2"))
wounded <-qs_read(here("output/data/seurat_objects/Wounded_7dpf_Level_02_immuneCells.qs2"))
wounded$Sample_Name <- factor(wounded$Sample_Name, levels = c("Control_kdrl", "Wounded_kdrl")) #wildtype always first
silva_immune_all <- qs_read(here("output/data/seurat_objects/Silva2021_28dpf_AllImmuneCells_juvenile.qs2"))
silva_MG <- qs_read(here("output/data/seurat_objects/Silva2021_28dpf_CAM_MGonly_juvenile.qs2"))


#general output path
output <- here("output/figures/")

```


## Colours 

```{r}
#celltype colours

celltype_colours <- c("red2", "black", "grey")
names(celltype_colours) <- c("MG/Macro subpop", "Neutrophil", "MG")

stage.colours <- c( '#e4c6d3','#be7794', '#a95175', '#823e59', '#3f1e2c', "#190c11")
names(stage.colours) <-  c('40hpf', '3dpf', '4dpf', '5dpf', '7dpf', "30dpf")

expression_colours <- c("#d9d9d9", "#40004b")

celltype_immune_colours <- rev(c("#544D51", 
                      "#D3CFD4", 
                      "#EB2C23", 
                      "#c3d5f7",
                      "#3876E6",
                      "#D667F8",
                      "#36BDBA",
                      "#F19D36", 
                      "#B5136D", 
                      "beige", 
                      "bisque1"))


names(celltype_immune_colours) <- levels(silva_immune_all$celltype)

wounded_colours <- c(
  "#dbe2c6", #wildtype
  "#657c95" #wounded
)

names(wounded_colours) <- c("Control_kdrl", "Wounded_kdrl")

```

## Genes

```{r}
genes <- c("apoc1", "apoeb", "p2ry12", "sall3a", #microglia
           "csf1rb", "spi1a", "mpeg1.1", "marco", #shared
           "cybb", "f13a1b", "cd74b", "tgfbi", #CAM
            "lyve1a","lyve1b", "mrc1a","mrc1b" , "dab2", "stab1", "stab2" , #functional genes for BAM equivalence
           "mpx", "mmp9", "lyz"#neutro
           )

genes_immune_annotation <- c("pcna", "mki67", "top2a", #prolif
                             "apoc1", "apoeb", "slco2b1", "sall3a", "p2ry12", "tgfbr1b", "slc2a5", #MG
                              "mpeg1.1","csf1ra", "csf1rb", #shared MG + CAM
                             "c1qa", "cybb", "marco", "f13a1b", "anxa5b", "nampta", "pla2g7", "tgfb1", "lyve1b", #CAM
                            "xcr1a.1", "flt3", "batf3", #DC
                            "cd79a", "cd79b", "pax5", #Bcell
                            "gzm3.3", "il2rb", "ccl38.6", #NK
                            "zap70", "lck", "il7r", #Tcell
                            "tbx21", "runx3", #ILC1
                            "il4", "gata3", "il13" #ILC2
                            )
```



# Wounded dataset panels


```{r, fig.align='center', fig.height=8, fig.width=15, warning=FALSE}
names(celltype_colours) <- c( "MG/Macro subpop\nn=70","Neutrophil\nn=27","MG\nn=252")
name <- "Wounded_GSE270486"
```


```{r, fig.align='center', fig.height=8, fig.width=15, warning=FALSE}

#use function
p1 <- plot_umap(seurat.object = wounded, 
                group = "myeloid_celltype", 
                cols = celltype_colours, 
                title = "Myeloid populations (n=349)", 
                project.name = name)
p2 <- plot_umap(seurat.object = wounded, 
                group = "Sample_Name", 
                cols = wounded_colours, 
                title = "Wounding status", 
                project.name = name)

#dotplot
p3 <- DotPlot(wounded, features = genes, group.by = "myeloid_celltype", cols = expression_colours) + rotate_x + ylab("") + ggtitle("Gene expression patterns") 
ggsave(sprintf("%s/%s_dotplot_myeloid_celltype.pdf", output, name),
         p3,
         device="pdf",
         width=20,
         height=10,
         units="cm",
         scale=1)

#stacked barplot
p4 <- wounded@meta.data %>% 
  group_by(myeloid_celltype, Sample_Name) %>% 
  summarise(., count =n()) %>% 
  ggplot(., aes(x = myeloid_celltype, y= count, fill=Sample_Name)) +
  geom_bar(stat="identity", position = "stack") + 
  scale_fill_manual(values = wounded_colours) +
  geom_text(aes(label = count), 
            position = position_stack(vjust=0.5),size=5, color="black")+
  plot_sizing  + xlab("")  +rotate_x + ggtitle("Wounded status composition")
ggsave(sprintf("%s/%s_stacked_barplot_myeloid_celltype_wounded.pdf", output, name),
         p4,
         device="pdf",
         width=13,
         height=10,
         units="cm",
         scale=1)

  
(p1+p2+plot_layout(ncol=2, axes="collect")) +p3+p4 
```


# Timecourse dataset panels


```{r}
name <- "Timecourse_GSE270486_GSE188342"
names(celltype_colours) <- c("MG/Macro subpop\nn=941", "Neutrophil\nn=1651", "MG\nn=133")
```


```{r, fig.align='center', fig.height=8, fig.width=18, eval=FALSE}
#use function
p1 <- plot_umap(seurat.object = timecourse, 
                group = "prelim_celltype", 
                cols = celltype_colours, 
                title = "Myeloid populations (n=2725)", 
                project.name = name)
p2 <- plot_umap(seurat.object = timecourse, 
                group = "Stage", 
                cols = stage.colours, 
                title = "Stage", 
                project.name = name)
p3 <- DotPlot(object = timecourse, features = genes, group.by = "prelim_celltype", cols = expression_colours)+ rotate_x + ylab("") + xlab("") + ggtitle("Gene expression patterns")
ggsave(sprintf("%s/%s_dotplot_myeloid_celltype.pdf", output, name),
         p3,
         device="pdf",
         width=20,
         height=10,
         units="cm",
         scale=1)


design_layout <- c("
                   AB
                   CX")


wrap_elements(p1 & theme(plot.title = element_text(size = 18, face = "italic"))) +
  wrap_elements(p2 & theme(plot.title = element_text(size = 18, face = "italic")))+ 
  wrap_elements(p3 & theme(plot.title = element_text(size = 18, face = "italic")))+
  plot_layout(design = design_layout)

```



# Silva et al Immune cells panels

Add N to celltype names + rename "CAMs" and microglia.

```{r}
silva_MG$celltype <- case_when(silva_MG$celltype == "CAM" ~ "MG/Macro subpop\nn=428",
                               silva_MG$celltype == "prolif MG" ~ "prolif MG\nn=406",
                               silva_MG$celltype == "MG" ~ "MG\nn=2141"
                               )

silva_MG$celltype_subclusters <- case_when(silva_MG$celltype_subclusters == "CAM" ~ "MG/Macro subpop\nn=422",
                               silva_MG$celltype_subclusters == "MG_1" ~ "MG_1\nn=1609",
                               silva_MG$celltype_subclusters == "MG_2" ~ "MG_2\nn=511",
                               silva_MG$celltype_subclusters == "MG_3" ~ "MG_3\nn=105",
                               silva_MG$celltype_subclusters == "prolif MG" ~ "prolif MG\nn=328"
                               )
```


```{r}
name_all <- "Silva_All_ImmuneCells"
name_MG <- "Silva_MG"

celltype_silva_colours <- c("#544D51", 
                      "#D3CFD4", 
                      "#EB2C23")

names(celltype_silva_colours) <- c("prolif MG\nn=406", 
                                   "MG\nn=2141", 
                                   "MG/Macro subpop\nn=428")

colours_celltype_subclusters <- c("#EB2C23", "#544D51", "grey90", "grey70", "grey45")
names(colours_celltype_subclusters) <- c("MG/Macro subpop\nn=422", "prolif MG\nn=328",
                                         "MG_1\nn=1609", "MG_2\nn=511", "MG_3\nn=105")
```



```{r}
p1 <- plot_umap(seurat.object = silva_immune_all, 
                group = "celltype", 
                cols = celltype_immune_colours, 
                title = "Immune cells (n=6666)", 
                project.name = name_all)

p2 <- DotPlot(object = silva_immune_all, features = genes_immune_annotation, 
              group.by = "celltype",
              cols = expression_colours) + rotate_x + ggtitle("Annotation dotplot")
ggsave(sprintf("%s/%s_dotplot_allimmunecells.pdf", output, name_all),
         p2,
         device="pdf",
         width=25,
         height=10,
         units="cm",
         scale=1)

p3 <- plot_umap(seurat.object = silva_MG, 
                group = "celltype", 
                cols = celltype_silva_colours, 
                title = "Microglia subpopulations (n=2975)", 
                project.name = name_MG)

p4 <- DotPlot(object = silva_MG, features = genes, 
              group.by = "celltype",
              cols = expression_colours) + rotate_x + ggtitle("Annotation dotplot")
ggsave(sprintf("%s/%s_dotplot_MG_subpop.pdf", output, name_MG),
         p4,
         device="pdf",
         width=20,
         height=10,
         units="cm",
         scale=1)

p5 <- plot_umap(seurat.object = silva_MG, 
                group = "celltype_subclusters", 
                cols = colours_celltype_subclusters, 
                title = "Microglia subpopulations (n=2975)", 
                project.name = name_MG)

p6 <- silva_MG@meta.data %>% 
  group_by(., celltype_subclusters, JMclusters) %>% 
  summarise(., count=n()) %>% 
  ggplot(., aes(x=celltype_subclusters, y=count, fill=JMclusters)) +
  geom_bar(stat="identity", position = "fill") +
  plot_sizing + rotate_x + xlab("") + ggtitle("Map to original clusters")

ggsave(sprintf("%s/%s_stacked_barplot_celltype_subclusters_OG_clusters.pdf", output, name_MG),
         p6,
         device="pdf",
         width=10,
         height=10,
         units="cm",
         scale=1)

p7 <- DotPlot(object = silva_MG, features = c("ctsa", "ctsba", "c1qa", "c1qc"), group.by = "celltype_subclusters", cols = expression_colours) + rotate_x + ggtitle("Location marker genes")
ggsave(sprintf("%s/%s_dotplot_locationmarker.pdf", output, name_MG),
         p7,
         device="pdf",
         width=12,
         height=8,
         units="cm",
         scale=1)

```


Make two figures

```{r, fig.align='center', fig.height=15, fig.width=15}

design_layout <- c("
                   AABBB
                   AABBB
                   CCDDD
                   CCDDD
                   EEFFF
                   EEFFF
                   GGFFF
                   GGXXX")


wrap_elements(p1 & theme(plot.title = element_text(size = 18, face = "italic"))) +
  wrap_elements(p2 & theme(plot.title = element_text(size = 18, face = "italic")))+ 
  wrap_elements(p3 & theme(plot.title = element_text(size = 18, face = "italic")))+
  wrap_elements(p4& theme(plot.title = element_text(size = 18, face = "italic")))+
  wrap_elements(p5& theme(plot.title = element_text(size = 18, face = "italic")))+
  wrap_elements(p6& theme(plot.title = element_text(size = 18, face = "italic"))) +
  wrap_elements(p7& theme(plot.title = element_text(size = 18, face = "italic")))+
  plot_layout(design = design_layout)


```


